---
title: "Index-based approaches"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Index-based approaches}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(offtake)
```

## The idea

Index-based methods ask a simple question: is the hunted site worse off than a
site that is not hunted (or only lightly hunted)? You measure the same thing at
both places, for example animal density or hunting yield, and compare them. If
the hunted site is clearly lower, that is a warning sign that hunting may not be
sustainable.

These methods are popular because they need little data and are easy to explain
(Adounke et al. 2026; Weinbaum et al. 2013). Their main limit is that they only
compare sites. They do not tell you the exact level of hunting at which things
tip over, and a difference between two sites can have causes other than hunting
(soil, habitat, or the size of the forest patch). So treat the verdict as a
signal to look closer, not as a final answer.

The package covers three of these indices:

| Function    | Compares                         | Warning sign                        |
|-------------|----------------------------------|-------------------------------------|
| `ot_pdc()`  | animal density                   | lower density at the hunted site    |
| `ot_hyco()` | hunting yield (or catch per unit effort) | lower yield at the more hunted site |
| `ot_asc()`  | share of young animals           | fewer juveniles at the hunted site  |

All three take a tidy data frame and unquoted column names, and all three return
a small table with a `sustainable` column (`TRUE` or `FALSE`).

We use the bundled example data `bushmeat_sites` (density and yield per
transect) and `bushmeat_ages` (one row per animal). Both are made up for the
sake of the examples, not real field data.

```{r}
head(bushmeat_sites)
```

## Population density comparison, `ot_pdc()`

Here you compare animal density at the hunted site with the reference site. You
give one row per density estimate (for example one row per transect), the
density column, the column that marks the site, and which site is the reference.

```{r}
ot_pdc(bushmeat_sites,
       density = density,
       group = site_type,
       reference = "reference")
```

How to read the output:

* `hunted_value` and `reference_value` are the mean densities at each site.
* `pct_change` is the change of the hunted site relative to the reference,

  $$\text{pct\_change} = \frac{\text{hunted} - \text{reference}}{\text{reference}} \times 100.$$

  A negative value means the hunted site holds fewer animals.
* `p_value` comes from a one-sided *t*-test asking whether the hunted site is
  really lower (it is `NA` if a site has fewer than two replicates).
* `sustainable` is `FALSE` when the hunted site is significantly lower, and
  `TRUE` otherwise.

## Hunting yield comparison, `ot_hyco()`

The logic is the same, but the metric is what hunters bring back rather than
what is left in the forest. If you also pass an `effort` column, the function
compares catch per unit effort (yield divided by effort) instead of raw yield.
That is often more honest, because more effort naturally brings more catch.

```{r}
ot_hyco(bushmeat_sites,
        yield = yield_kg,
        group = site_type,
        reference = "reference",
        effort = effort_days)
```

The output columns match `ot_pdc()`. Here `hunted_value` and `reference_value`
hold the mean catch per hunter-day at each site.

## Age structure comparison, `ot_asc()`

This one looks at who is being caught. In a healthy population you expect a fair
share of young animals. When hunting is heavy, juveniles often become scarce,
which is a sign that the population is not replacing itself. So a lower share of
juveniles at the hunted site is the warning sign.

You give one row per animal, the class column, the site column, and which class
counts as juvenile.

```{r}
ot_asc(bushmeat_ages,
       stage = age_class,
       group = site_type,
       reference = "reference",
       juvenile = "juvenile")
```

Here `juv_prop_hunted` and `juv_prop_reference` are the juvenile shares (between
0 and 1) at each site, and `pct_change` is the change relative to the reference.

If you prefer to compare the whole age structure at once rather than just the
juvenile share, use `method = "distribution"`. That runs a chi-squared test on
the full class by site table and flags a significant difference. In that case
read the result together with the direction of the shift, since a difference on
its own does not say which way things are going.

```{r}
ot_asc(bushmeat_ages,
       stage = age_class,
       group = site_type,
       reference = "reference",
       method = "distribution")
```

## A note of caution

These indices are quick screens. A `FALSE` verdict says the hunted site looks
worse on this one measure, which is worth following up. It does not prove that
hunting is the cause. Whenever you can, make sure the two sites are otherwise
comparable, and read the three indices together rather than trusting any single
one.

## References

Adounke, G. R. M. et al. (2026) Systematic review of sustainability assessment
approaches for wildlife exploitation. *Biological Conservation* 313, 111606.

Weinbaum, K. Z., Brashares, J. S., Golden, C. D. & Getz, W. M. (2013) Searching
for sustainability: are assessments of wildlife harvests behind the times?
*Ecology Letters* 16, 99 to 111.
