---
title: "Model-based approaches"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Model-based approaches}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(offtake)
```

## The idea

Model-based methods work the other way round from the index-based ones. Instead
of comparing two sites, they use the biology of the species to estimate how many
animals can be taken each year without shrinking the population. You then
compare that estimate with the offtake you actually observe. If the observed
offtake is larger than the estimate, hunting is likely too heavy (Adounke et al.
2026).

Each function returns a small table with a `sustainable` column that is `TRUE`
when the observed harvest sits at or below the estimated safe level.

| Function     | Estimates the safe harvest from            | Source                    |
|--------------|--------------------------------------------|---------------------------|
| `ot_pro()`   | density, growth rate, lifespan             | Robinson & Redford 1991   |
| `ot_pbr()`   | a cautious population size and growth rate  | Wade 1998                 |
| `ot_msy()`   | the logistic growth curve                   | Schaefer 1954             |
| `ot_samse()` | growth rate plus year-to-year variability   | Manlik et al. 2022        |
| `ot_biode()` | settlement locations and hunting effort     | Levi et al. 2009          |

We use the bundled example data `duiker_demography` (life-history values for a
few duikers) and `manu_settlements` (hunting villages on a map). Both are made
up for the examples.

```{r}
duiker_demography
```

## Production model, `ot_pro()`

This is the most common model in bushmeat studies. The safe harvest, called the
production $P$, is

$$P = 0.6 \, K \, (\lambda_{\max} - 1) \, F.$$

Reading it piece by piece:

* $K$ is the density the habitat can support (carrying capacity).
* $0.6\,K$ is the density at which the population grows fastest, so it produces
  the most surplus.
* $\lambda_{\max}$ is the fastest yearly growth the species can manage. The part
  $(\lambda_{\max} - 1)$ is the surplus it adds each year.
* $F$ is a safety fraction that depends on how long the animal lives:
  0.6 for short-lived species (about 5 years), 0.4 for medium-lived (5 to 10
  years) and 0.2 for long-lived species (over 10 years). Long-lived animals
  bounce back slowly, so you take a smaller share.

If you do not have `lambda_max`, you can let the package work it out from three
life-history numbers using `ot_lambda_max()`: the number of female young per
female per year (`b`), the age at first reproduction (`a`) and the age at last
reproduction (`w`). This solves Cole's (1954) equation, a standard way to turn
basic life-history into a growth rate.

```{r}
ot_lambda_max(b = 0.55, a = 1, w = 10)
```

Now the model itself. Here we pass the life-history columns, so `lambda_max` is
computed for us, and we use `lifespan` to set the safety fraction `F`.

```{r}
ot_pro(duiker_demography,
       k = density_k,
       harvest = annual_take,
       b = b, a = a, w = w,
       longevity = lifespan)
```

The `production` column is the estimated safe harvest, `harvest` is what was
observed, and `sustainable` compares the two.

## Potential biological removal, `ot_pbr()`

`ot_pbr()` comes from marine mammal management and is deliberately cautious. The
safe removal is

$$\mathrm{PBR} = N_{\min} \times \tfrac{1}{2} R_{\max} \times F_R,$$

where $N_{\min}$ is a low, cautious estimate of population size (so you do not
over-count), $R_{\max}$ is the maximum yearly growth rate, and $F_R$ is a
recovery factor between 0.1 and 1 that you lower when you want to be extra
careful. If you give a point estimate `n` and its coefficient of variation `cv`,
the function works out $N_{\min}$ for you as the lower end of the plausible
range.

```{r}
ot_pbr(duiker_demography,
       rmax = rmax,
       removal = annual_take,
       n = density_k,
       fr = 0.5)
```

## Maximum sustainable yield, `ot_msy()`

This is the classic fisheries benchmark. It assumes the population follows the
logistic curve, where growth is fastest at half of carrying capacity and slows
as the population fills up. The largest steady catch it can support is

$$\mathrm{MSY} = \frac{rK}{4},$$

with $r$ the growth rate and $K$ the carrying capacity. If you also give the
current abundance `n`, the function reports the surplus at that abundance too,
which is the sustainable yield right now rather than at the ideal population
size.

```{r}
ot_msy(duiker_demography,
       r = rmax,
       k = density_k,
       harvest = annual_take)
```

## Stochastic safe mortality, `ot_samse()`

Real environments are not steady. Good years and bad years come and go, and that
variability eats into how much you can safely take. `ot_samse()` follows this
idea, after Manlik et al. (2022): it projects the population many times with
random good and bad years, taking a fixed number $H$ each year,

$$N_{t+1} = \max\!\left(0,\; N_t\, e^{r_t} - H\right), \qquad r_t \sim \mathcal{N}(r_{\max}, \sigma_e),$$

and searches for the largest $H$ that still keeps the population from shrinking
on average. Here $r_{\max}$ is the mean growth rate and $\sigma_e$ is how much it
swings from year to year.

Because it uses simulation, you supply the mean growth rate (`rmax`), how much it
varies from year to year (`sd_env`), the observed removal, and a starting
population size. The result is usually a bit lower than what a formula without
variability would suggest, which is the point.

```{r}
# add a variability value and a starting size to the example data
samse_data <- transform(duiker_demography,
                        sd_env = 0.2,
                        n0 = density_k)

ot_samse(samse_data,
         rmax = rmax,
         sd_env = sd_env,
         removal = annual_take,
         n0 = n0,
         nsims = 200,
         years = 40,
         seed = 1)
```

This is a simplified version of the original method, which was built in the
Vortex software and includes more biological detail. Use it as a careful screen,
and cite Manlik et al. (2022) for the full approach.

## Spatial source-sink model, `ot_biode()`

The models above treat the landscape as one pot. In practice hunting is not
spread evenly: it is heavy near villages and light far away. `ot_biode()`, after
Levi et al. (2009), captures this. Near a settlement the animals get depleted
(a sink), while remote areas stay full and quietly resupply the hunted zones (a
source).

You give the location of each settlement and how many hunters live there, plus a
few parameters for the species and the hunting. The model then predicts animal
density at every point of the map,

$$N(x, y) = \left[K^{\theta}\left(1 - \frac{e\,d\,h}{r}\sum_c P_c\,\varphi(\delta_c)\right)\right]_+^{1/\theta},
\qquad \varphi(\delta) = \frac{\exp\!\left(-\delta^2 / 2\sigma^2\right)}{2\pi\delta + 1}.$$

In words: each settlement $c$ presses down on the animals around it. That
pressure grows with its hunter population $P_c$ and fades with distance $\delta_c$
through the kernel $\varphi$ (how far hunters roam is set by $\sigma$). Add up the
pressure from all settlements, and what the population can still support is
whatever is left of its carrying capacity $K$. The $[\,\cdot\,]_+$ just means the
density cannot go below zero. You do not need to handle any of this by hand; the
function does it across a grid.

```{r}
res <- ot_biode(manu_settlements,
                x = x_km,
                y = y_km,
                humans = hunters,
                k = 25,
                r = 0.07,
                hphy = 40,
                kill_rate = 0.1,
                sigma = 6,
                resolution = 2)
res
```

The summary reports, among other things, `frac_extirpated` (the share of the map
pushed to local extinction) and `mean_density`. The full predicted surface is
available with `ot_biode_surface()`, ready for mapping, and the catch per unit
effort each village can expect is in `ot_biode_cpue()`.

```{r}
head(ot_biode_surface(res))
ot_biode_cpue(res)
```

If you have ggplot2 installed, the surface maps nicely as a raster, with the
depletion halos around the settlements.

```{r, eval = requireNamespace("ggplot2", quietly = TRUE), fig.width = 6, fig.height = 4, fig.alt = "Map of predicted game density across the landscape, showing lower density around each hunting settlement and higher density far from settlements."}
library(ggplot2)

surface <- ot_biode_surface(res)

ggplot(surface, aes(x, y, fill = density)) +
  geom_raster() +
  geom_point(data = manu_settlements,
             aes(x_km, y_km), inherit.aes = FALSE,
             colour = "white", shape = 4, size = 2) +
  scale_fill_viridis_c(name = "density") +
  coord_equal() +
  labs(title = "Predicted game density", x = "km", y = "km") +
  theme_minimal()
```

## Which one to use

There is no single best model. The production model is a good default when you
have density and basic life-history. PBR is handy when you mostly trust a
population count and want to stay cautious. MSY is familiar from fisheries.
SAMSE is worth reaching for when year-to-year swings matter. Biode is the one to
use when the spatial pattern of hunting is the whole point. In practice it helps
to try more than one and see whether they agree.

## References

Adounke, G. R. M. et al. (2026) Systematic review of sustainability assessment
approaches for wildlife exploitation. *Biological Conservation* 313, 111606.

Robinson, J. G. & Redford, K. H. (1991) Sustainable harvest of Neotropical
forest mammals.

Wade, P. R. (1998) Calculating limits to the allowable human-caused mortality of
cetaceans and pinnipeds. *Marine Mammal Science* 14, 1 to 37.

Levi, T., Shepard, G. H., Ohl-Schacherer, J., Peres, C. A. & Yu, D. W. (2009)
Modelling the long-term sustainability of indigenous hunting in Manu National
Park, Peru. *Journal of Applied Ecology* 46, 804 to 814.

Manlik, O. et al. (2022) A stochastic model for estimating sustainable limits to
wildlife mortality in a changing world. *Conservation Biology* 36, e13897.
